---
title: "Assignment 4 Task 2: Parameter estimation"
author: "Grace Brofman"
date: "3/4/2021"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Introduction:**

Data on global wild fish catch and aqucaulture production was compiled by the Earth Policy Institute (EPI) and U.N. Food and Agriculture Organization (FAO) from 1950-2010. Here, nonlinear least squares (NLS) is used to estimate parameters for global wild fish catch.

Below: exploratory graphs of wild catch over time (both as-is (top) and log-transformed (bottom))
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(nlstools)
library(janitor)
library(lubridate)
library(tsibble)

fish_catch <- read_csv("fish_catch.csv") %>% 
  janitor::row_to_names(2) %>% 
  clean_names() %>% 
  rename(wild_catch_mill_tons = wild_catch)

fish_clean <- fish_catch[-c(1,2, 66, 67, 68, 69),] # removing rows 1 and 2 - need extra comma!

index_vector <- c(0:62) # creating vector to add as column to data frame

fish_clean$index_vector <- index_vector # adding vector as data frame

# converting to numeric for graphing:
fish_clean$wild_catch_mill_tons <- as.numeric(fish_clean$wild_catch_mill_tons)
fish_clean$year <- as.Date(fish_clean$year, "%Y") # classifying as date (year)

# explorator graph of wild fish catch over time
# logistic growth
exploratory_graph <- ggplot(data = fish_clean,
       aes(x = year,
           y = wild_catch_mill_tons)) +
  geom_point() +
  labs(x = "year", y = "wild catch (millions of tons")
exploratory_graph

# log transformed
log_explore <- ggplot(data = fish_clean,
       aes(x = year,
           y = log(wild_catch_mill_tons))) +
  geom_point() +
  labs(x = "year", y = "ln(wild catch(millions of tons))")
log_explore

```
```{r}
# Exploratory graph appears somewhat logistic, Log-transformed data looks somewhat smoothed out. The later in time we get, the more of an upper limit fish catch appears to reach, indicating a carrying capacity. It looks similar to logistic growth, although the left "leg" of the graph does not appear to completely follow logistic growth.
```

Exploratory graphs suggest a logistic relationship.

Logistic growth equation: 

$P(t)=\frac{K}{1+Ae^{-kt}}$, where

- $K$ is the carrying capacity
- $A$ is $\frac{K-P_0}{P_0}$
- $k$ is the growth rate constant

Parameter estimates using nonlinear least squares (NLS):

- $K$ ~ 94 million tons 
- $A$ ~ 4.52
- $k$ ~ 0.057 million tons/year


```{r, message = FALSE, warning = FALSE, include = FALSE}
# First, estimate the growth constant during the exponential growth phase (*k*)

# not removing observations here because growth appears to be exponential immediately
df_exp <- fish_clean %>% 
  filter(index_vector < 21) %>% 
  mutate(ln_catch = log(wild_catch_mill_tons))

# now model linear growth to get *k* estimate (aka slope of linear equation)
lm_k <- lm(ln_catch ~ index_vector, data = df_exp)
lm_k

# coefficient (k) ~ 0.057

```

```{r, message = FALSE, warning = FALSE, include = FALSE}
# Estimating parameters using non-linear least squares (NLS):
df_nls <- nls(wild_catch_mill_tons ~ K/(1 + A*exp(-r*index_vector)),
              data = fish_clean,
              start = list(K = 94, A = 4.52, r = 0.057),
              trace = TRUE)

# see the model summary (null hypothesis: parameter value = 0)
summary(df_nls)
```
```{r, message = FALSE, warning = FALSE}
# use broom:: functions to get model outputs in tidier format:
model_out <- broom::tidy(df_nls) # use View(model_out) to see the tidy model output
View(model_out)


```

Our model with estimated parameters is:
$$P(t) = \frac{100.3}{1+4.32e^{-0.07t}}$$

```{r, message = FALSE, warning = FALSE}
# make predictions for catch at all times in orginal df:
catch_predict <- predict(df_nls)

# bind predictions to original df:
df_complete <- data.frame(fish_clean, catch_predict)

# plot them together:
ggplot(data = df_complete,
       aes(x = year, y = wild_catch_mill_tons)) +
  geom_point() +
  geom_line(aes(x = year, y = catch_predict),
            color = "blue") +
  theme_minimal() + 
  labs(x = "Year", y = "Wild catch (millions of tons)", 
       title = "Actual vs. predicted global wild fish catch (1950 - 2012)") +
  geom_annotate()

```

### **Source:** 

Global wild fish catch and aquaculture production, compiled by Earth Policy Institute with 1950-2010 from U.N. Food and Agriculture Organization (FAO), Global Capture Production and Global Aquaculture Production, electronic databases, at www.fao.org/fishery/topic/16140/en.
